\vspace{-.5cm} Lights: a generalized joint model for high-dimensional multivariate longitudinal data and censored durations \vspace{.5cm}

### 2021-08-27 paper.tex
Update missing index
Update missing time interval in baseline hazard formula
Add one more tidle g function
Update matrix notation
remove MC_sep option idea in the paper for now

Given~\eqref{eq:link-function} and Assumption~\ref{indep-hyp-2}, each $y_i^l$ is assumed to be from a one-parameter exponential family with respect to a reference measure which is either the Lebesgue measure (e.g., in the Gaussian case) or the counting measure (e.g., in the logistic cases).
The conditional distribution of $y_i|b_i, G_i=k$ is then assumed to be from a distribution with a density of the form
\begin{equation}
	\label{eq:exp-family-density}
	f(y_i|b_i, G_i=k) = \exp \big\{(y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \big\},
\end{equation}
with 
\[ \Phi_i = (\phi_1^{-1} {\textbf{1}_{n_i^1}}^\top \cdots \phi_L^{-1} {\textbf{1}_{n_i^L}}^\top)^\top \in \R^{n_i} \] 
and $\phi = (\phi_1, \ldots, \phi_L)^\top \in \R^L$.
The density described in~\eqref{eq:exp-family-density} encompasses several distributions, see Table~\ref{table:glm}. The functions $c_\phi(\cdot)$ and $d_\phi(\cdot)$ are known as well as the dispersion parameters $\phi_l$, while parameters $\beta_k$ have to be estimated. Note that $d_\phi(\cdot)$ is related to the normalizing constant.
\begin{table}[htb]
\centering
\resizebox{\textwidth}{!}{%
\begin{tabular}{cccccc}
\toprule
 Model & Support & Use cases & $\phi_l$ & $h_l(\cdot)$ in~\eqref{eq:link-function} & $c_\phi(\cdot)$ \\
 \midrule
 Gaussian & $\R$ & Continuous response data & $\sigma_l^2$ & $z \mapsto z$ & $z \mapsto z^2 / 2 \sigma_l^2$ \\ 
 Categorical & $\{0, 1\}^C$ & Outcome with $C$ modalities & 1 & $z \mapsto \log\Big(\dfrac{z}{1 - z} \Big)$ & $z \mapsto \log(1 + e^z)$ \\
 Poisson & $\N$ & Count of occurrences & 1 &  $z \mapsto \log(z)$ & $z \mapsto e^z$ \\
 \bottomrule
\end{tabular}}
\caption[]{Examples of standard distributions that fit in the considered setting, given in the univariate case for simplicity of the notations.}
\label{table:glm}
\end{table}

@@

\noindent where $\Sigma_i$ is a diagonal matrix whose diagonal is $\Phi_i$.

\noindent One can either choose $\left\{ \begin{array}{ll} \text{M}_i^1 &= W_i \\ \text{M}_{ik}^2 &= W_iV_i^T\Sigma_i^{-1}(y_i - U_i\beta_k) \end{array} \right.$ based on~\eqref{eq:Lambda_ik-b_i|y_i}, where in the Gaussian case used in practice, one can sample from
\begin{align*}
  y_i | G_i = k &\sim \cN\big(U_i\beta_k, V_iDV_i^\top + \Sigma_i\big) \\
  b_i | y_i, G_i=k &\sim \cN \big(W_iV_i^T\Sigma_i^{-1}(y_i - U_i\beta_k), W_i\big)  
\end{align*}
with $W_i = (V_i^\top\Sigma_i^{-1}V_i + D^{-1})^{-1}$, or one can choose $\left\{ \begin{array}{ll} \text{M}_i^1 &= D^{(w)} \\ \text{M}_{ik}^2 &= 0 \end{array} \right.$ based on~\eqref{eq:Lambda_ik-b_i} using a sampling $b_i \sim \cN(0, D^{(w)})$. The advantage of this second option is that $S_{ik}^{(w)}$ does not depend on $i$ nor $k$, which significantly improves computational performances, even though the variance is bigger than the first option (see Appendix~\ref{sec:?}). We then let the practitioner choose the option in our implementation, since the second one could be more appropriated for a cross-validation phase for instance.

@@

\vspace{.5cm}
Lemma~\ref{lemma:prox-sgl1} bellow states that one can easily compute the proximal operator of the sparse group lasso defined in Algorithm~\ref{alg:ISTA}.
\begin{lemma}
\label{lemma:prox-sgl1}
The proximal operator of the sparse group lasso can be expressed as the composition of the group lasso and the lasso proximal operators, that is
\normalfont
\begin{equation*}
  \text{prox}_{\text{sg} l_1,\tilde \eta, \zeta} = \text{prox}_{\zeta \tilde \eta \sum_{l=1}^L \norm{\cdot}_{2,l}} \circ \text{prox}_{\zeta (1 - \tilde \eta) \norm{\cdot}_1}.
\end{equation*}
\end{lemma}
The two proximal operators on the right hand side of Lemma~\ref{lemma:prox-sgl1} are both well known analytically~\citep{bach2011optimization} and tractable. We used the \texttt{Python} library \texttt{copt} for the implementation of proximal gradient descent~\citep{copt} and we propose in our package an implementation for the proximal operator of the sparse group lasso based on Lemma~\ref{lemma:prox-sgl1}.
A proof of Lemma~\ref{lemma:prox-sgl1} is proposed in Appendix~\ref{sec:proof-lemma-prox-sgl1}.
Hence, we update $\gamma_k^{(w)}$ with the following step
\begin{equation}
  \label{eq:gamma-update}
  \gamma_k^{(w+1)} = \text{ISTA}_{\text{sg} l_1}\big(\gamma_k^{(w)}, Q^{(w)}_{n,k}(\cdot), \tilde \eta, \zeta_{2,k}\big)
\end{equation}
Note that the warmstart initialization is left to the user's choice, as well as the ISTA acceleration~\citep{beck2009fast}, since warmstart can diminish acceleration effectiveness~\citep{tibshirani2010proximal}. The gradient of $Q_{n,k}^{(w)}$ is here given by
\begin{align}
  \label{eq:grad-Q}
  &\frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_a(t_i, b_i^l) | t_i, \delta_i, y_i] \\ \nonumber
   &\qquad\qquad - \sum_{j=1}^{J-1} (t_{j+1}^u-t_j^u) \lambda_0^{(w)}(t_j^u) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j, b_i^l) \exp \big\{\gamma_k^\top \Psi_i(t_j^u)\big\} | t_i, \delta_i, y_i] \ind{\{t_j^u < t_i\}} \Big]
\end{align}
for all $a = 1, \ldots, \mathcal{A}$ and $l = 1, \ldots, L$.

Then we consider a piecewise-constant estimator $\lambda_0^{(w)}$ for the baseline hazard with jumps at each uncensored observed event times. Hence, we treat the jump size as parameters to be estimated~\citep{klein1992semiparametric}, and the closed-form update obtained by cancelling the gradient is given by
\begin{equation}
  \label{eq:update_lambda_0}
  \lambda_0^{(w+1)}(t_j)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t_j=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \hat \E_{\theta^{(w)}}\Big[\exp \big\{{\gamma_k^{(w+1)}}^\top \Psi_i(t_j)\big\} | t_i, \delta_i, y_i \Big] \ind{\{t_i \geq t_j\}} },
\end{equation}
for all $j=1, \ldots, n$, which is a Breslow like estimator~\citep{breslow1972contribution} adapted to our model.

@@

Let us introduce the following useful notations
\begin{equation}
  \label{eq:g-tilde}
  \left\{
    \begin{array}{ll}
    \tilde g^1&: b_i \mapsto b_i,\\
    \tilde g^2&: b_i \mapsto b_ib_i^\top, \\
    \tilde g^3_{s}&: b_i \mapsto \Psi_i(s), \\
    \tilde g^4_{s, \gamma_k}&: b_i \mapsto \exp \big\{\gamma_k^\top g^3_{s}(b_i) \big\}, \\
    \tilde g^5_{s, \gamma_k}&: b_i \mapsto g^3_{s}(b_i)^\top g^4_{s}(b_i).
    \end{array}
  \right. 
\end{equation}
We also denote 
\[ \cI_1 = 
\begin{bmatrix}
  \ind{\{t_{1} = t_1^u\}}  & \cdots & \ind{\{t_{1} = t_J^u\}}\\
  \vdots &  \ddots & \vdots \\
  \ind{\{t_{n} = t_1^u\}} & \cdots & \ind{\{t_{n} = t_J^u\}}
\end{bmatrix}
\in \R^{n \times J}
\quad \text{and} \quad 
\cI_2 = 
\begin{bmatrix}
  \ind{\{t_{1} \leq t_1^u\}} & \cdots & \ind{\{t_{1} \leq t_J^u\}}\\
  \vdots &  \ddots & \vdots \\
  \ind{\{t_{n} \leq t_1^u\}} & \cdots & \ind{\{t_{n} \leq t_J^u\}}
\end{bmatrix}
\in \R^{n \times J} \]
some matrices to compare each event times with the vector of distinct event times. Those matrices are widely used in practice to replace indicators. For instance
\begin{align*}
f(t_i, \delta_i| b_i, G_i = k ; \theta^{(w)}) &= \big[\lambda_0^{(w)}(t_i) \exp \big\{{\gamma_k^{(w)}}^\top\Psi_i(t_i) \big\} \big]^{\delta_i} \exp \Big( - \sum_{j=1}^{J-1} (t_{j+1}^u-t_j^u) \lambda_0^{(w)}(t_j^u) \exp \big\{{\gamma_k^{(w)}}^\top \Psi_i(t_j^u)\big\} \ind{\{t_j^u < t_i\}} \Big) \\
& = \big[\lambda_0^{(w)}(t_i) g^4_{t_i, \gamma_k}(b_i) \big]^{\delta_i} \exp \Big( - \sum_{j=1}^{J-1} (t_{j+1}^u-t_j^u) \lambda_0^{(w)}(t_j^u) g^4_{t_j^u, \gamma_k}(b_i) \ind{\{t_j^u < t_i\}} \Big)
\end{align*}

\texttt{TODO Van Tuan : update matrix formulation.}

>>>>

Given~\eqref{eq:link-function} and Assumption~\ref{indep-hyp-2}, each $y_i^l$ is assumed to be from a one-parameter exponential family with respect to a reference measure which is either the Lebesgue measure (e.g., in the Gaussian case) or the counting measure (e.g., in the logistic cases).
The conditional distribution of $y_i|b_i, G_i=k$ is then assumed to be from a distribution with a density of the form
\begin{equation}
	\label{eq:exp-family-density}
	f(y_i|b_i, G_i=k) = \exp \big\{(y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \big\},
\end{equation}
with 
\begin{equation}
  \label{eq:def-Phi_i} 
  \Phi_i = (\phi_1^{-1} {\textbf{1}_{n_i^1}}^\top \cdots \phi_L^{-1} {\textbf{1}_{n_i^L}}^\top)^\top \in \R^{n_i}
\end{equation} 
and $\phi = (\phi_1, \ldots, \phi_L)^\top \in \R^L$.
The density described in~\eqref{eq:exp-family-density} encompasses several distributions, see Table~\ref{table:glm}. The functions $c_\phi(\cdot)$ and $d_\phi(\cdot)$ are known as well as the dispersion parameters $\phi_l$, while parameters $\beta_k$ have to be estimated. Note that $d_\phi(\cdot)$ is related to the normalizing constant.
\begin{table}[htb]
\centering
\resizebox{\textwidth}{!}{%
\begin{tabular}{cccccc}
\toprule
 Model & Support & Use cases & $\phi_l$ & $h_l(\cdot)$ in~\eqref{eq:link-function} & $c_\phi(\cdot)$ \\
 \midrule
 Gaussian & $\R$ & Continuous response data & $\sigma_l^2$ & $z \mapsto z$ & $z \mapsto z^2 / 2 \sigma_l^2$ \\ 
 Categorical & $\{0, 1\}^C$ & Outcome with $C$ modalities & 1 & $z \mapsto \log\Big(\dfrac{z}{1 - z} \Big)$ & $z \mapsto \log(1 + e^z)$ \\
 Poisson & $\N$ & Count of occurrences & 1 &  $z \mapsto \log(z)$ & $z \mapsto e^z$ \\
 \bottomrule
\end{tabular}}
\caption[]{Examples of standard distributions that fit in the considered setting, given in the univariate case for simplicity of the notations.}
\label{table:glm}
\end{table}

@@

\noindent One can either choose $\left\{ \begin{array}{ll} \text{M}_i^1 &= W_i \\ \text{M}_{ik}^2 &= W_iV_i^T\Sigma_i^{-1}(y_i - U_i\beta_k) \end{array} \right.$ based on~\eqref{eq:Lambda_ik-b_i|y_i} and where $\Sigma_i$ is the diagonal matrix whose diagonal is $\Phi_i$ (defined in~\eqref{eq:def-Phi_i}), where in the Gaussian case used in practice, one can sample from
\begin{align*}
  y_i | G_i = k &\sim \cN\big(U_i\beta_k, V_iDV_i^\top + \Sigma_i\big) \\
  b_i | y_i, G_i=k &\sim \cN \big(W_iV_i^T\Sigma_i^{-1}(y_i - U_i\beta_k), W_i\big)  
\end{align*}
with $W_i = (V_i^\top\Sigma_i^{-1}V_i + D^{-1})^{-1}$, or one can choose $\left\{ \begin{array}{ll} \text{M}_i^1 &= D^{(w)} \\ \text{M}_{ik}^2 &= 0 \end{array} \right.$ based on~\eqref{eq:Lambda_ik-b_i} using a sampling $b_i \sim \cN(0, D^{(w)})$. The advantage of this second option is that $S_{ik}^{(w)}$ does not depend on $i$ nor $k$, which significantly improves computational performances. We then implement this option in practice, being more appropriated in our high-dimensional context.

@@

\vspace{.5cm}
Lemma~\ref{lemma:prox-sgl1} bellow states that one can easily compute the proximal operator of the sparse group lasso defined in Algorithm~\ref{alg:ISTA}.
\begin{lemma}
\label{lemma:prox-sgl1}
The proximal operator of the sparse group lasso can be expressed as the composition of the group lasso and the lasso proximal operators, that is
\normalfont
\begin{equation*}
  \text{prox}_{\text{sg} l_1,\tilde \eta, \zeta} = \text{prox}_{\zeta \tilde \eta \sum_{l=1}^L \norm{\cdot}_{2,l}} \circ \text{prox}_{\zeta (1 - \tilde \eta) \norm{\cdot}_1}.
\end{equation*}
\end{lemma}
The two proximal operators on the right hand side of Lemma~\ref{lemma:prox-sgl1} are both well known analytically~\citep{bach2011optimization} and tractable. We used the \texttt{Python} library \texttt{copt} for the implementation of proximal gradient descent~\citep{copt} and we propose in our package an implementation for the proximal operator of the sparse group lasso based on Lemma~\ref{lemma:prox-sgl1}.
A proof of Lemma~\ref{lemma:prox-sgl1} is proposed in Appendix~\ref{sec:proof-lemma-prox-sgl1}.
Hence, we update $\gamma_k^{(w)}$ with the following step
\begin{equation}
  \label{eq:gamma-update}
  \gamma_k^{(w+1)} = \text{ISTA}_{\text{sg} l_1}\big(\gamma_k^{(w)}, Q^{(w)}_{n,k}(\cdot), \tilde \eta, \zeta_{2,k}\big)
\end{equation}
Note that the warmstart initialization is left to the user's choice, as well as the ISTA acceleration~\citep{beck2009fast}, since warmstart can diminish acceleration effectiveness~\citep{tibshirani2010proximal}. The gradient of $Q_{n,k}^{(w)}$ is here given by
\begin{align}
  \label{eq:grad-Q}
  &\frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_a(t_i, b_i^l) | t_i, \delta_i, y_i] \\ \nonumber
   &\qquad\qquad - \sum_{j=1}^{J-1} (t_{j+1}^u-t_j^u) \lambda_0^{(w)}(t_j^u) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j^u, b_i^l) \exp \big\{\gamma_k^\top \Psi_i(t_j^u)\big\} | t_i, \delta_i, y_i] \ind{\{t_j^u < t_i\}} \Big]
\end{align}
for all $a = 1, \ldots, \mathcal{A}$ and $l = 1, \ldots, L$.

Then we consider a piecewise-constant estimator $\lambda_0^{(w)}$ for the baseline hazard with jumps at each uncensored observed event times. Hence, we treat the jump size as parameters to be estimated~\citep{klein1992semiparametric}, and the closed-form update obtained by cancelling the gradient is given by
\begin{equation}
  \label{eq:update_lambda_0}
  \lambda_0^{(w+1)}(t_j)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t_j=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} (t_{j+1}^u-t_j^u) \hat \E_{\theta^{(w)}}\Big[\exp \big\{{\gamma_k^{(w+1)}}^\top \Psi_i(t_j)\big\} | t_i, \delta_i, y_i \Big] \ind{\{t_i \geq t_j\}} },
\end{equation}
for all $j=1, \ldots, n$, which is a Breslow like estimator~\citep{breslow1972contribution} adapted to our model.

@@

Let us introduce the following useful notations
\begin{equation}
  \label{eq:g-tilde}
  \left\{
    \begin{array}{ll}
    \tilde g^1&: b_i \mapsto b_i,\\
    \tilde g^2&: b_i \mapsto b_ib_i^\top, \\
    \tilde g^3_{s}&: b_i \mapsto \Psi_i(s), \\
    \tilde g^4_{s, \gamma_k}&: b_i \mapsto \exp \big\{\gamma_k^\top g^3_{s}(b_i) \big\}, \\
    \tilde g^5_{s, \gamma_k}&: b_i \mapsto g^3_{s}(b_i)^\top g^4_{s}(b_i).
    \end{array}
  \right. 
\end{equation}
We also denote 
\[ \cI_1 = 
\begin{bmatrix}
  \ind{\{t_{1} = t_1^u\}}  & \cdots & \ind{\{t_{1} = t_J^u\}}\\
  \vdots &  \ddots & \vdots \\
  \ind{\{t_{n} = t_1^u\}} & \cdots & \ind{\{t_{n} = t_J^u\}}
\end{bmatrix}
\in \R^{n \times J}
\quad \text{and} \quad 
\cI_2 = 
\begin{bmatrix}
  \ind{\{t_{1} \leq t_1^u\}} & \cdots & \ind{\{t_{1} \leq t_J^u\}}\\
  \vdots &  \ddots & \vdots \\
  \ind{\{t_{n} \leq t_1^u\}} & \cdots & \ind{\{t_{n} \leq t_J^u\}}
\end{bmatrix}
\in \R^{n \times J} \]
some matrices to compare each event times with the vector of distinct event times. Those matrices are widely used in practice to replace indicators. For instance
\begin{align*}
f(t_i, \delta_i| b_i, G_i = k ; \theta^{(w)}) &= \big[\lambda_0^{(w)}(t_i) \exp \big\{{\gamma_k^{(w)}}^\top\Psi_i(t_i) \big\} \big]^{\delta_i} \exp \Big( - \sum_{j=1}^{J-1} (t_{j+1}^u-t_j^u) \lambda_0^{(w)}(t_j^u) \exp \big\{{\gamma_k^{(w)}}^\top \Psi_i(t_j^u)\big\} \ind{\{t_j^u < t_i\}} \Big) \\
& = \big[\lambda_0^{(w)}(t_i) g^4_{t_i, \gamma_k}(b_i) \big]^{\delta_i} \exp \Big( - \sum_{j=1}^{J-1} (t_{j+1}^u-t_j^u) \lambda_0^{(w)}(t_j^u) g^4_{t_j^u, \gamma_k}(b_i) \ind{\{t_j^u < t_i\}} \Big) \\
f(T, \Delta| S^{(w)}, G = k ; \theta^{(w)}) &= \big\{ \underset{J}{\text{sum }} \big[g^4_{\gamma_k} \lambda_0^{(w)} \cI_1 \big] \big\}^{\Delta} \exp \Big( - \underset{J}{\text{sum }} \big[g^4_{\gamma_k} \lambda_0^{(w)} \Delta T\cI_2 \big] \Big)
\end{align*}
where $\Delta T = (t^u_2 - t^u_1  \cdots t^u_J - t^u_{J - 1}) \in \R^{J - 1}$.
